Hippocampal replay of experience at real-world speeds

### 2020-10-08 Article.tex
Update on Overleaf.
Draft3

Using a 0.8 threshold, we were able to classify 89\% (23382 of 26160) of SWRs as containing at least one of the three dynamics or common dynamic mixtures. To ensure that this depended on the sequential firing of spatially tuned cells in the hippocampus, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. We then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs were classified in the shuffled datasets, a value that was significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, Figure \ref{fig:Figure3-Figure supplement 2}).

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods). We found that 96\% (22478 of 23382) of classified SWRs and 85\% of all SWRs included spatially coherent structure and the prevalence of spatially coherent structure was consistent across animals (Figure \ref{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We found that only 3353 of 23382 or 14\% of classified SWRs had time points with spatially incoherent structure (Figure \ref{3}G).

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4511 of 23382 or 19\% of classified SWRs had time periods where the decoded position was classified as continuous (Figure \ref{3}G, 17\% of all SWRs). Thus, focusing on only high speed trajectories misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

We repeated our classification analysis with a higher classification threshold of 0.95 to ensure that our result was not dependent on the threshold of 0.8. We found that, while this change slightly reduced the total fraction of classified SWRs (19180 / 25845 or 74\% of SWRs), an even higher fraction of the classified SWRs (19020 / 19180 or 99\% classified) had spatially coherent content. Similarly, SWRs containing spatially incoherent content were a small fraction of the classified SWRs (spatially incoherent: 484 / 19180 or 3\%).

@@

We found that spatially coherent events did indeed have lower cumulative HPD sizes than events with fragmented dynamics. Figure \ref{4}A and D show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The most probable positions (as estimated by the model) at each time step in these SWRs is concentrated in a small portion of the track and correspondingly, the cumulative HPD size is small throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the most probable positions are more spatially diffuse and the cumulative HPD size is much larger. The HPD size also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}B, where the cumulative HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the average cumulative HPD size for each SWR was much smaller than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). Additionally, the HPD size for the unclassified SWRs was similarly large. We note here that while the size of the HPD will be influenced by the dynamic estimated from the data, it remains the case that a continuous or stationary dynamic can correspond to a large HPD region (see outliers in Figure \ref{4}E), indicating less certainty for those specific events.  To show this, as before, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes had much larger HPD size than the real data (epoch \#1: 55 cm vs. 229 cm, p=0.02, real vs. shuffled; epoch \#2: 72 cm vs. 231 cm, p=0.02, real vs. shuffled; one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).

\begin{figure*}%[tb2]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v6}
\caption{Validation of classification using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. In each column: top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. Middle panel: Posterior distribution of estimated linear position over time. Magenta dashed line indicates animal's position. Bottom panel: Cumulative HPD size---the integral of the position bins with the top 95\% probable values. \textbf{(E)} Average Cumulative 95\% HPD size for each dynamic category. Median 24 cm for spatially coherent vs. median 238 cm for spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U}
\label{4}
\end{figure*}

@@

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, and thus correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (16454 of 23382 classified SWRs or 70\%) are well described by a single dynamic or mixtures of dynamics (Figure \ref{5}A). Surprisingly, the most common category is stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at slower speeds (Figure \ref{5}D; though note these speeds are calculated from the MAP estimate and thus noisy), and are slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position and cannot be explained by erroneously detected SWRs.

The second most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these events represented a location close to the animal's actual position (Figure \ref{5}C). There were, however, a number of stationary events that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify non-local stationary events and found such sequences in 1551 of 23382 (~7\%) of classified SWRs. About half of these (731 classified SWRs) were stationary throughout the duration of the SWR. These non-local stationary events most commonly represented reward wells or choice points (Figure \ref{5}E), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}, but a small portion of occurred at other locations of the track. This suggests that these representations can be flexibly deployed to represent discrete locations.

@@

Previous work reported that less than half of SWRs or high-population-activity events contained replay \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset, the large majority of SWRs (85\%) contained spatially coherent content, as defined by some combination of stationary or continuous dynamics. These events included decoded representations where the most probable positions were highly spatially concentrated. This was in stark contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

@@

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead, there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. These representation do not move far in the span of a SWR. Interestingly, a previous paper \cite{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. We believe that the difference in these findings is a result of defining "chance" levels of occurrence of these events, using large time bins, and using a restrictive decoding model. We also note that we chose to include a stationary distribution to explicitly evaluate the prevalence of stationary events in adult animals, but in principle a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds close to 0 m/s.

@@

\begin{contributions}

\end{contributions}

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% of values of the posterior probability \cite{CasellaStatisticalinference2001}. In this manuscript, we use the HPD to evaluate the uncertainty of the posterior probability of position. We calculate the 95\% HPD by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The cumulative spatial extent of the HPD is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Riemann sum.

@@

\subsection*{Estimation of Speed}
In order to estimate the speed over time, we first found the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is defined as the center of the position bin with the greatest posterior value. We then computed the first derivative using the Python library Numpy's gradient function, which computes the difference in the forward and backward time direction and then averages them for all points except the boundaries. We then smoothed this quantity with a small Gaussian (2.5 ms standard deviation) and then averaged this for the classified time points.

@@

%TC:endignore
%the command above ignores this section for word count

>>>>

Using a 0.8 threshold, we were able to classify 89\% (23,382 of 26,160) of SWRs as containing at least one of the three dynamics or common dynamic mixtures. To ensure that this depended on the sequential firing of spatially tuned cells in the hippocampus, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. We then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs were classified in the shuffled datasets, a value that was significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, Figure \ref{fig:Figure3-Figure supplement 2}).

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. We therefore asked what fraction of classified events contain spatially coherent content, which we define as SWRs containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods). We found that 96\% (22,478 of 23,382) of classified SWRs and 85\% of all SWRs included spatially coherent structure and the prevalence of spatially coherent structure was consistent across animals (Figure \ref{3}G). We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented or fragmented-continuous mixture dynamics. We found that only 3353 of 23,382 or 14\% of classified SWRs had time points with spatially incoherent structure (Figure \ref{3}G).

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4511 of 23,382 or 19\% of classified SWRs had time periods where the decoded position was classified as continuous (Figure \ref{3}G, 17\% of all SWRs). Thus, focusing on only high speed trajectories misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

We repeated our classification analysis with a higher classification threshold of 0.95 to ensure that our result was not dependent on the threshold of 0.8. We found that, while this change slightly reduced the total fraction of classified SWRs (19,180 of 25,845 or 74\% of SWRs), an even higher fraction of the classified SWRs (19,020 of 19,180 or 99\% classified SWRs) had spatially coherent content. Similarly, SWRs containing spatially incoherent content were a small fraction of the classified SWRs (spatially incoherent: 484 of 19,180 or 3\% classified SWRs).

@@

We found that spatially coherent events did indeed have lower HPD sizes than events with fragmented dynamics. Figure \ref{4}A and D show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The most probable positions (as estimated by the model) at each time step in these SWRs is concentrated in a small portion of the track and correspondingly, the HPD size is small throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the most probable positions are more spatially diffuse and the HPD size is much larger. The HPD size also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}B, where the HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the average HPD size for each SWR was much smaller than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). Additionally, the HPD size for the unclassified SWRs was similarly large. We note here that while the size of the HPD will be influenced by the dynamic estimated from the data, it remains the case that a continuous or stationary dynamic can correspond to a large HPD region (see outliers in Figure \ref{4}E), indicating less certainty for those specific events. In general, if the data does not exhibit strong place specific firing, the HPD size will be large, no matter the influence of the dynamics. To show this, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes had much larger HPD size than the real data (epoch \#1: 55 cm vs. 229 cm, p=0.02, real vs. shuffled; epoch \#2: 72 cm vs. 231 cm, p=0.02, real vs. shuffled; one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).

\begin{figure*}%[tb2]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v6}
\caption{Validation of classification using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. In each column: top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. Middle panel: Posterior distribution of estimated linear position over time. Magenta dashed line indicates animal's position. Bottom panel: HPD size---the integral of the position bins with the top 95\% probable values. \textbf{(E)} Average 95\% HPD size for each dynamic category. Median 24 cm for spatially coherent vs. median 238 cm for spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U}
\label{4}
\end{figure*}

@@

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, and thus correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (16454 of 23382 or 70\% classified SWRs) are well described by a single dynamic or mixtures of dynamics (Figure \ref{5}A). Surprisingly, the most common category is stationary-continuous mixtures (9860 of 23382 or 42\% classified SWRs, Figure \ref{5}A, Figure\ref{fig:Figure5-Figure supplement 1}). These events contain representations that move at slower speeds (Figure \ref{5}D; though note these speeds are calculated from the MAP estimate and thus noisy when the estimated position is more uncertain i.e. has a large HPD size), and are slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position and cannot be explained by erroneously detected SWRs.

The second most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\% classified SWRs). Unlike the stationary-continuous mixtures, most of these events represented a location close to the animal's actual position (Figure \ref{5}C). There were, however, a number of stationary events that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify non-local stationary events and found such sequences in 1551 of 23382 (~7\%) of classified SWRs. About half of these (731 classified SWRs) were stationary throughout the duration of the SWR. These non-local stationary events most commonly represented reward wells or choice points (Figure \ref{5}E), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}, but a small portion of occurred at other locations of the track. This suggests that these representations can be flexibly deployed to represent discrete locations.

@@

Previous work reported that less than half of SWRs or high-population-activity events contained replay \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset, the large majority of SWRs (85\%) contained spatially coherent content, as defined by some combination of stationary or continuous dynamics. These events included decoded representations where the most probable positions were highly spatially concentrated (i.e. had small HPD size). This was in stark contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

@@

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead, there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. Interestingly, a previous paper \cite{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. We believe that the difference in these findings is a result of defining "chance" levels of occurrence of these events, using large time bins, and using a restrictive decoding model. We also note that we chose to include a stationary distribution to explicitly evaluate the prevalence of stationary events in adult animals, but in principle a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds close to 0 m/s.

@@

\begin{contributions}
E.L.D. conceived of the model and did the analysis with help from U.T.E, A.K.G., and M.E.C. E.L.D and L.M.F wrote the manuscript with help from A.K.G. and M.E.C. A.K.G, K.K., M.S., J.E.C. contributed data to this manuscript.
\end{contributions}

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% of values of the posterior probability \cite{CasellaStatisticalinference2001}. In this manuscript, we use the HPD to evaluate the uncertainty of the posterior probability of position. We calculate the 95\% HPD by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The HPD size is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Riemann sum.

@@

\subsection*{Estimation of Speed}
In order to estimate the speed over time, we first found the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is defined as the center of the position bin with the greatest posterior value. We then computed the first derivative using the Python library Numpy's gradient function, which computes the difference in the forward and backward time direction and then averages them for all points except the boundaries. We then smoothed this quantity with a small Gaussian (2.5 ms standard deviation) and then averaged this for the classified time points. Because our position bin size of 3 cm makes it hard to distinguish between slower speeds in a 2 ms time step, we only analyzed classifications longer than 20 ms.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5-supplemental1/Figure5_v1_supplemental1}
\caption{More examples of stationary-continuous-mixtures.}
\label{fig:Figure5-Figure supplement 1}
\end{figure*}
%TC:endignore
%the command above ignores this section for word count
